---
layout: page
title: Programming with R
subtitle: Loading Data
minutes: 30
output: 
  html_document: 
    keep_md: yes
---

> ## Learning Objectives {.objectives}
> * Read tabular data from a file into a program.
> * Assign values to variables.
> * Select individual values and subsections from data.
> * Perform operations on a data frame of data.


The data sets are stored in [comma-separated values](reference.html#comma-separated-values-(csv)) (CSV) format. 
The first few rows of our first file look like this:

```{r echo = FALSE}
tmp <- read.csv("data/sample.csv", header = FALSE, nrows = 5)
write.table(tmp, quote = FALSE, sep = ",", row.names = FALSE, col.names = FALSE)
rm(tmp)
```

We want to:

* Load data into memory,
* Look at different data types and how to manipulate data
* Plot some results

To do all that, we'll have to learn a little bit about programming.

### Loading Data

Now that we have our data files, we need to start RStudio. Since we've pre-made a project, the easiest way to do this is to double click on the `SWC_R_exercise.r` file inside our project. RStudio should automatically put us into the correct directory, and show us the files available under the `Files` tab. However, we can also check in the Console:


```{r}
getwd()
```

Just like in the Unix Shell, we type the command and then press `Enter` (or `return`).
getwd() shows us the working directory. If we hadn't set up a project beforehand, we could also specifically set a working directory:

```{r, eval=FALSE}
setwd("C:/Users/Amanda/Desktop/SWC_R_exercise")
```

But since we're already in that directory, this command has no effect. Now we can load the data into R using `read.csv`:

```{r echo = FALSE, eval=TRUE}
tmp <- read.csv("data/sample.csv")
write.table(tmp, quote = FALSE, sep = ",", row.names = FALSE, col.names = FALSE)
rm(tmp)
```

```{r eval=FALSE}
read.csv(file = "data/sample.csv")
```

The expression `read.csv(...)` is a [function call](reference.html#function-call) that asks R to run the function `read.csv`.

`read.csv` has two [arguments](reference.html#argument): the name of the file we want to read, and whether the first line of the file contains names for the columns of data.
The filename needs to be a character string (or [string](reference.html#string) for short), so we put it in quotes. Assigning the second argument, `header`, to be `FALSE` indicates that the data file does not have column headers. We'll talk more about the value `FALSE`, and its converse `TRUE`, in lesson 04.

> ## Tip {.callout}
>
> `read.csv` actually has many more arguments that you may find useful when
> importing your own data in the future. You can learn more about these
> options in this supplementary [lesson](01-supp-read-write-csv.html).

The utility of a function is that it will perform its given action on whatever value is passed to the named argument(s).
For example, in this case if we provided the name of a different file to the argument `file`, `read.csv` would read it instead.
We'll learn more of the details about functions and their arguments in the next lesson.

Since we didn't tell it to do anything else with the function's output, the console will display the full contents of the file `sample.csv`.
Try it out.

`read.csv` read the file, but we can't use data unless we assign it to a variable.
A variable is just a name for a value, such as `x`, `current_temperature`, or `subject_id`.
We can create a new variable simply by assigning a value to it using `<-`


```{r}
weight_kg <- 55
```

Once a variable has a value, we can print it by typing the name of the variable and hitting `Enter` (or `return`).
In general, R will print to the console any object returned by a function or operation *unless* we assign it to a variable.

```{r}
weight_kg
```

We can do arithmetic with the variable:

```{r}
# weight in pounds:
2.2 * weight_kg
```

> ## Tip {.callout}
>
> We can add comments to our code using the `#` character. It is useful to
> document our code in this way so that others (and us the next time we
> read it) have an easier time following what the code is doing.

We can also change an object's value by assigning it a new value:

```{r}
weight_kg <- 57.5
# weight in kilograms is now
weight_kg
```

If we imagine the variable as a sticky note with a name written on it,
assignment is like putting the sticky note on a particular value:

<img src="fig/python-sticky-note-variables-01.svg" alt="Variables as Sticky Notes" />

This means that assigning a value to one object does not change the values of other variables.
For example, let's store the subject's weight in pounds in a variable:

```{r}
weight_lb <- 2.2 * weight_kg
# weight in kg...
weight_kg
# ...and in pounds
weight_lb
```

<img src="fig/python-sticky-note-variables-02.svg" alt="Creating Another Variable" />

and then change `weight_kg`:

```{r}
weight_kg <- 100.0
# weight in kg now...
weight_kg
# ...and weight in pounds still
weight_lb
```

<img src="fig/python-sticky-note-variables-03.svg" alt="Updating a Variable" />

Since `weight_lb` doesn't "remember" where its value came from, it isn't automatically updated when `weight_kg` changes.
This is different from the way spreadsheets work.

> ## Tip {.callout} 
>An alternative way to print the value of a variable is to use () around the assignment statement. As an example: `(total_weight <- weight_kg + weight_lb)`, adds the values of `weight_kg` and `weight_lb`, assigns the result to the `total_weight`, and finally prints the assigned value of the variable `total_weight`.


Now that we know how to assign things to variables, let's re-run `read.csv` and save its result:

```{r}
aneurism <- read.csv(file = "data/sample.csv")
```

This statement doesn't produce any output because assignment doesn't display anything.
If we want to check that our data has been loaded, we can print the variable's value.
However, for large data sets it is convenient to use the function `head` to display only the first few rows of data.

```{r}
head(aneurism)
```

> ## Challenge - Assigning values to variables {.challenge}
>
> Draw diagrams showing what variables refer to what values after each statement in the following program:
>
```{r}
mass <- 47.5
age <- 122
mass <- mass * 2.0
age <- age - 20
```

### Manipulating Data

Now that our data is loaded in memory, we can start doing things with it.
First, let's ask what type of thing `aneurism` is:

```{r}
class(aneurism)
```

The output tells us that is a data frame. Think of this structure as a spreadsheet in MS Excel that many of us are familiar with.
Data frames are very useful for storing data and you will find them elsewhere when programming in R. A typical data frame of experimental data contains individual observations in rows and variables in columns.

We can see the dimensions, or [shape](reference.html#shape-(of-an-array)), of the data frame with the function `dim`:

```{r}
dim(aneurism)
```

This tells us that our data frame, `aneurism`, has `r nrow(aneurism)` rows and `r ncol(aneurism)` columns.

If we want to get a single value from the data frame, we can provide an [index](reference.html#index) in square brackets, just as we do in math:

```{r}
# first value in aneurism
aneurism[1, 1]
# middle value in aneurism
aneurism[30, 20]
```

An index like `[30, 20]` selects a single element of a data frame, but we can select whole sections as well.
For example, we can select the first ten days (columns) of values for the first four patients (rows) like this:

```{r}
aneurism[30:40, 1:3]
```

The [slice](reference.html#slice) `1:4` means, "Start at index 1 and go to index 4."

The slice does not need to start at 1, e.g. the line below selects rows 5 through 10:

```{r}
aneurism[1:5, 6:9]
```
We can use the function `c`, which stands for **c**ombine, to select non-contiguous values:

```{r}
aneurism[c(1, 30, 54, 72), c(2, 4, 5)]
```

We also don't have to provide a slice for either the rows or the columns.
If we don't include a slice for the rows, R returns all the rows; if we don't include a slice for the columns, R returns all the columns.
If we don't provide a slice for either rows or columns, e.g. `aneurism[, ]`, R returns the full data frame.

```{r}
# All columns from row 5
aneurism[5, ]
# All rows from column 16
aneurism[, 6]
```

### Adding Elements

The function `c()` (for combine) can also be used to add elements to a vector.

```{r}
z <- 1:10
z <- c(z, 11)
z
z <- c(0, z)
z
```

> ## Tip {.callout} 
>An alternative way to get numbers in a sequence is to use `seq()`. Try the expression `seq(from = 1, to = 10, by = 0.1)` 


### Addressing by Name

Columns in an R data frame are named.

```{r column_names}
colnames(aneurism)
```

> ## Tip {.callout} 
>
> If names are not specified e.g. using `headers=FALSE` in a `read.csv()` function, R assigns default names `V1,V2,...,Vn`

We usually use the `$` operator to address a column by name

```{r named_addressing}
aneurism$Gender
```

Named addressing can also be used in square brackets.
```{r names_addressing_2}
head(aneurism[ , c('Age','Gender')])
```

> ## Best Practice {.callout} 
>
> Best practice is to address columns by name, often you will create or delete columns and the column position will change.


We can use logical vectors to select data from a data frame.

```{r logical_vectors_indexing}
index <- aneurism$Group == 'Control'
aneurism[index,]$BloodPressure
```

Often this operation is written as one line of code:

```{r logical_vectors_indexing2}
aneurism[aneurism$Group=='Control',]$BloodPressure
```

> ## Challenge - Using logical indexes {.challenge}
> 1. Create a scatterplot showing BloodPressure for subjects not in the control group.
> 2. How many ways are there to index this set of subjects?


We are studying aneurism patients. Each row holds the observations for just one patient.

Now let's perform some common mathematical operations to learn about our data.
When analyzing data we often want to look at partial statistics, such as the maximum value per patient or the average value per day.
One way to do this is to select the data we want to create a new temporary data frame, and then perform the calculation on this subset:

```{r}
# first row, all of the columns
BloodP <- aneurism$BloodPressure
# max inflammation for patient 1
max(BloodP)
```

We don't actually need to store the row in a variable of its own.
Instead, we can combine the selection and the function call:

```{r}
# max inflammation for patient 2
max(aneurism$BloodPressure)
```

R also has functions for other common calculations, e.g. finding the minimum, mean, median, and standard deviation of the data:

```{r}
# minimum patient age
min(aneurism$Age)
# mean patient age
mean(aneurism$Age)
# median patient age
median(aneurism$Age)
# standard deviation of patient age
sd(aneurism$Ag)
```
